---
title: "Differentially private synthesis (Track B)"
output:
  litedown::html_format:
    meta:
      css: ["@default"]
vignette: >
  %\VignetteEngine{litedown::vignette}
  %\VignetteIndexEntry{Differentially private synthesis (Track B)}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
library(flexsynth)
```

## Two tracks, one interface

`flexsynth` has two engines. **Track A** (the default) maximises statistical
utility and carries *no* formal privacy guarantee — you judge the residual risk
with `diagnose()` and `disclosure_risk()`. **Track B** trades some utility for a
mathematically provable guarantee: **differential privacy (DP)**. You opt in by
passing a `dp_control()` to the `privacy` argument of `synth()`; nothing else
about the call changes.

```{r optin}
dp <- dp_control(epsilon = 1, delta = 1e-6, mechanism = "gaussian")
dp
```

Differential privacy bounds how much *any single individual* can change the
output distribution. The guarantee is stated as a pair
(\(\epsilon\), \(\delta\)): smaller values mean stronger privacy and more noise.
The default privacy unit here is **person-level** — the guarantee protects an
individual across *all* of their rows, not just a single record.

## A worked example

We use a small **synthetic** cardiac dataset (never real patient data): one row
per patient, with an age, sex, a smoking flag and a systolic blood pressure that
depends on both.

```{r data}
n <- 1500
real <- data.frame(
  id     = seq_len(n),
  age    = round(rnorm(n, 62, 11)),
  sex    = factor(sample(c("F", "M"), n, TRUE, prob = c(0.45, 0.55))),
  smoker = sample(c(FALSE, TRUE), n, TRUE, prob = c(0.7, 0.3))
)
real$sbp <- round(0.6 * real$age + ifelse(real$smoker, 8, 0) + rnorm(n, 90, 10))
head(real)
```

### Where the bin edges come from

The DP engine works over a **discrete grid**: numeric variables are cut into
bins. Where those bin edges come from matters, because reading them straight from
the data (its min and max) can betray an individual's presence — an outlier's
exact value *is* the maximum. The `domain` argument controls this, and the
default keeps the accounting honest without any work on your part:

- `domain = "dp"` (default): numeric variables you name in `bounds` use those
  public edges for free; any others have their working range **estimated under
  differential privacy** and the cost is folded into the reported
  (\(\epsilon\), \(\delta\)). Nothing leaks unaccounted.
- `domain = "public"`: every numeric variable *must* have a public range in
  `bounds` (an error otherwise); no budget is spent on the domain. Use this when
  you have codebook or physiological limits.
- `domain = "data"`: the old behaviour — edges read from the data with a warning,
  *excluded* from the accounting. Benchmarking only.

Categorical columns carry their domain in their type, so pass them as `factor`
(or `logical`) — their levels are public metadata. A bare `character` column is
refused in the rigorous modes.

For a clean, fully data-independent release with **no** budget spent on the
domain, pass public ranges via `bounds`:

```{r bounds}
dp <- dp_control(
  epsilon    = 2,
  delta      = 1e-6,
  mechanism  = "gaussian",
  bounds     = list(age = c(18, 100), sbp = c(60, 240))
)
res <- synth(real, structure = ~ id, privacy = dp, seed = 1)
res
```

The result is an ordinary `synth_result`; `as.data.frame()` gives the synthetic
table, whose `id` is a fresh surrogate key.

```{r syn}
syn <- as.data.frame(res)
head(syn)
```

### The privacy accounting

Every DP result carries an accounting record describing exactly what was spent
and how the noise was calibrated. This is the object to keep with a governed
release.

```{r acct}
res$privacy
```

The engine measures a set of low-order marginals under a single, composed budget:
one-way marginals for every variable and (with the default
`dependence = "tree"`) the pairwise marginals needed to learn a Chow-Liu
dependency tree. The number of marginals, the per-cell noise scale, and — for the
Gaussian mechanism — the zero-concentrated-DP \(\rho\) are all reported.

### Or let the engine estimate them privately

If you have no public ranges, the default `domain = "dp"` estimates each numeric
variable's working range under DP — a clamp-free exponential-mechanism quantile
at each end — and charges it to the budget (a small slice, `domain_frac`, default
10%). No warning, and the reported (\(\epsilon\), \(\delta\)) is exact:

```{r estimate}
dp_auto <- dp_control(epsilon = 2, delta = 1e-6, mechanism = "gaussian")
res_auto <- synth(real, structure = ~ id, privacy = dp_auto, seed = 1)
res_auto$privacy
```

The accounting names the variables whose edges were estimated and the fraction of
budget it took. The remaining budget pays for the marginals, exactly as before.

## Choices that matter

**Mechanism.** `"laplace"` gives *pure* \(\epsilon\)-DP (set `delta = 0`);
`"gaussian"` gives approximate (\(\epsilon\), \(\delta\))-DP with tighter
composition when many marginals are measured (it needs `delta > 0`).

**Dependence.** `"tree"` (default) keeps first- and second-order structure by
fitting a Chow-Liu tree over the noisy pairwise marginals — no extra budget,
because the same measurements are reused for structure and parameters.
`"independent"` measures only one-way marginals: fewer queries, so less noise per
query, but cross-variable correlations are lost. The real `age`/`sbp`
correlation here is about `r round(cor(real$age, real$sbp), 2)`. Under DP the tree
recovers a substantial part of it while the independent model necessarily drives
it to zero:

```{r dependence}
b <- list(age = c(18, 100), sbp = c(60, 240))
dp_tree  <- dp_control(epsilon = 6, mechanism = "laplace", dependence = "tree",
                       bins = 8, bounds = b)
dp_indep <- dp_control(epsilon = 6, mechanism = "laplace", dependence = "independent",
                       bins = 8, bounds = b)
s_tree  <- as.data.frame(synth(real, ~ id, privacy = dp_tree,  seed = 1))
s_indep <- as.data.frame(synth(real, ~ id, privacy = dp_indep, seed = 1))
c(real  = cor(real$age, real$sbp),
  tree  = cor(s_tree$age,  s_tree$sbp),
  indep = cor(s_indep$age, s_indep$sbp))
```

Two things are visible in that vector: the tree keeps a clear positive
association (never the full strength — DP noise attenuates it), and the
independent model has no correlation at all by construction. The gap between the
tree and the real value is the honest price of the guarantee; a larger `epsilon`
or a coarser `bins` narrows it.

**Budget-efficient structure learning.** With `dependence = "tree"` the default
fitter measures *every* pairwise marginal at full fidelity, then keeps only the
\(d-1\) that end up in the Chow-Liu tree — so budget was spent measuring pairs the
model discards. `structure_frac` splits the work instead: a small fraction of the
marginal budget buys a deliberately rough all-pairs scan used *only* to pick the
tree, and the rest is concentrated on re-measuring just the chosen edges. Because
structure *selection* tolerates noise far better than the conditional *parameters*
do, this sharpens the surviving edges — increasingly so as the number of variables
grows (with only a handful it is roughly break-even, and it is inert below three).
Both passes are reported and compose into the *same* exact budget — the cheap scan
and the concentrated re-measurement are two sequential releases whose
pure-\(\epsilon\) adds (or, for Gaussian, whose zCDP \(\rho\) adds):

```{r structure-frac}
dp_eff <- dp_control(epsilon = 6, mechanism = "laplace", dependence = "tree",
                     bins = 8, bounds = b, structure_frac = 0.25)
synth(real, ~ id, privacy = dp_eff, seed = 1)$privacy
```

The accounting line splits into the pairwise scans and the parameter marginals,
and the per-cell noise is reported separately for each — the scan is noisier (it
only has to rank dependencies), the parameters sharper.

**Contribution bound.** At `unit = "person"` each person may contribute at most
`max_rows_per_person` rows (default `1`, appropriate for one-row-per-person
tables). If a person legitimately has several rows, set this from public domain
knowledge; rows beyond the cap are subsampled away before any budget is spent, so
the guarantee holds by construction.

**More budget, more fidelity.** Larger `epsilon` means less noise. The synthetic
marginals converge to the real ones as \(\epsilon\) grows:

```{r budget}
for (e in c(0.5, 2, 8)) {
  d <- dp_control(epsilon = e, mechanism = "laplace",
                  dependence = "independent",
                  bounds = list(age = c(18, 100), sbp = c(60, 240)))
  s <- as.data.frame(synth(real, ~ id, privacy = d, seed = 1))
  cat(sprintf("epsilon = %-3g  mean(sbp): real %.1f  syn %.1f\n",
              e, mean(real$sbp), mean(s$sbp)))
}
```

## Longitudinal releases: a DP Markov model

If the `structure` declares a nesting index (`~ id / visit`), Track B preserves
within-unit temporal structure with a first-order Markov model — the private
analogue of Track A's initial-state + lag-1 transition model. Three things are
measured under one composed budget: a **length histogram** (how many rows a
person contributes), the **initial-state marginals** (the `t = 1` row, one- and
two-way as before), and a **transition matrix** \(P(v_t \mid v_{t-1})\) for each
variable. Generation draws a length, an initial row, then steps each variable's
transition matrix, so autocorrelation across visits is retained.

Because a person now contributes a whole trajectory, you must set
`max_rows_per_person` to the public maximum number of visits — this bounds each
person's effect on the transition histograms (a length-\(\le c\) trajectory has
at most \(c - 1\) consecutive pairs). The budget still composes to exactly
(\(\epsilon\), \(\delta\)); the accounting reports the split.

```{r longitudinal}
long <- do.call(rbind, lapply(1:400, function(i) {
  nv <- sample(2:4, 1); s <- numeric(nv); s[1] <- rnorm(1, 130, 12)
  for (t in seq_len(nv)[-1]) s[t] <- 0.85 * s[t - 1] + 0.15 * 130 + rnorm(1, 0, 5)
  data.frame(id = i, visit = seq_len(nv), sbp = round(s),
             sex = sample(c("F", "M"), 1))
}))
long$sex <- factor(long$sex)

dp_long <- dp_control(epsilon = 8, mechanism = "laplace",
                      max_rows_per_person = 4, bounds = list(sbp = c(60, 240)))
res_long <- synth(long, structure = ~ id / visit, privacy = dp_long, seed = 1)
res_long$privacy
```

The synthetic `visit` index is regenerated as the within-person position, and a
lag-1 correlation like the real one survives the noise (DP attenuates it):

```{r longitudinal-cor}
lag1 <- function(d) {
  d <- d[order(d$id, d$visit), ]
  prev <- ave(d$sbp, d$id, FUN = function(x) c(NA, head(x, -1)))
  ok <- !is.na(prev); cor(prev[ok], d$sbp[ok])
}
syn_long <- as.data.frame(res_long)
c(real = lag1(long), synthetic = lag1(syn_long))
```

### Baseline columns held exactly constant

By default every non-index column is treated as time-varying, so a genuinely
subject-invariant covariate (here `sex`) is stepped through its own transition
matrix and can drift between a synthetic person's visits. If you know a column is
constant within a person — public schema knowledge, not something read from the
data — name it in `baseline`. It is then modelled **once** in the initial-state
model (keeping its distribution and its correlation with the first visit) and
broadcast unchanged to every row, so it is exactly constant within each synthetic
unit. Because a baseline column needs no transition histogram, declaring it also
*removes* that histogram from the release, so the remaining measurements are
sharper at the same (\(\epsilon\), \(\delta\)):

```{r baseline}
res_base <- synth(long, structure = ~ id / visit,
                  privacy = dp_control(epsilon = 8, mechanism = "laplace",
                                       max_rows_per_person = 4,
                                       bounds = list(sbp = c(60, 240)),
                                       baseline = "sex"),
                  seed = 1)

# every synthetic person now has a single sex across their visits
syn_base <- as.data.frame(res_base)
max(tapply(syn_base$sex, syn_base$id, function(s) length(unique(s))))
res_base$privacy   # one fewer transition histogram than the release above
```

### Higher-order and cross-variable transitions

The default transition model is first-order and per-variable: each variable's next
value depends only on its *own* previous value. Two knobs deepen it. `transition_order
= k` conditions on a variable's own last `k` values (momentum), and `transition_cross
= m` additionally conditions on the lag-1 values of its `m` most strongly associated
companions — so cross-variable coupling is re-measured at every step, not just carried
at `t = 1` by the initial-state tree. Companions are picked from the pairwise marginals
the tree already measures, so cross-conditioning spends **no** extra budget (a
transition tuple still lands in exactly one cell); a higher order actually *lowers* the
transition sensitivity to `cap - order`, since a person then contributes fewer, deeper
tuples. The order must therefore be at most `max_rows_per_person - 1`, and
`transition_cross > 0` needs the tree model.

```{r transition-order}
res_ho <- synth(long, structure = ~ id / visit,
                privacy = dp_control(epsilon = 8, mechanism = "laplace",
                                     dependence = "tree", max_rows_per_person = 4,
                                     bounds = list(sbp = c(60, 240)),
                                     baseline = "sex",
                                     transition_order = 2, transition_cross = 1),
                seed = 1)
res_ho$privacy   # note: transitions order 2 + 1 cross-parent, same (eps, delta)
```

Early rows (before position `order + 1`, which have no full history yet) are generated
by marginalising the very same measured tensor, which is post-processing and costs
nothing. The trade-off is cell sparsity: a wider conditioning grid spreads the same
noisy counts more thinly, so reach for depth only when the extra structure is worth it.

## Linked multi-table releases

`synth_linked()` accepts a `dp_control()` too, giving a DP release across a whole
key hierarchy at once. The privacy unit is the **root entity** (e.g. a patient):
adding or removing one individual — its root row *and* every descendant row that
cascades from it — changes the release within the budget. Contribution is bounded
hierarchically, so `max_rows_per_person` here means the maximum children kept per
parent for each child table: a single integer for every child table, or a named
list keyed by table name. The root cap is always 1.

```{r linked}
set.seed(1)
patients <- data.frame(
  id  = 1:300,
  age = round(rnorm(300, 60, 11)),
  sex = factor(sample(c("F", "M"), 300, TRUE)))
adm <- do.call(rbind, lapply(patients$id, function(pid) {
  n <- rpois(1, 1.4); if (n == 0) return(NULL)
  data.frame(id = pid, admission_id = seq_len(n), los = 1L + rpois(n, 4))
}))

dp_link <- dp_control(
  epsilon = 4, mechanism = "laplace",
  max_rows_per_person = list(admissions = 6),        # <= 6 admissions per patient
  domain = "public",
  bounds = list(age = c(18, 100), los = c(0, 60)))

res_link <- synth_linked(
  tables     = list(patients = patients, admissions = adm),
  structures = list(patients = ~ id, admissions = ~ id / admission_id),
  keys       = list(patients = "id", admissions = c("id", "admission_id")),
  privacy    = dp_link, seed = 1)
res_link$privacy
```

Each table's own variable marginals and a children-per-parent **count histogram**
are measured under one exactly-composed budget; the summed L1 (Laplace) and summed
squared L2 (Gaussian zCDP) sensitivities add over every histogram, with each
table's contribution scaled by its per-entity *path cap* (the product of branching
caps from the root). Synthetic children copy their synthetic parent's surrogate
key, so referential integrity holds by construction — `check_linkage()` confirms
it:

```{r linked-ri}
check_linkage(res_link)
```

By default each child table's variables are modelled by their own within-table
marginals, so the synthetic child *links* to a synthetic parent but is
statistically independent of it.

### Conditioning children on the synthetic parent

`dp_control(cross_table = TRUE)` closes that gap. For each child table with a
modellable immediate parent, the engine also measures **parent-by-child joint
marginals** — counted at the child grain (one observation per child row, at the
parent's value carried down the foreign key). Their person-sensitivity is the
child's path cap, exactly like a child one-way marginal, so they fold into the
same composed budget. The parent's variables then enter the child's Chow-Liu
structure as fixed context nodes, and at generation the synthetic parent's
already-drawn value conditions the child draw:

```{r linked-cross}
dp_cross <- dp_control(
  epsilon = 4, mechanism = "laplace", cross_table = TRUE,
  max_rows_per_person = list(admissions = 6),
  domain = "public",
  bounds = list(age = c(18, 100), los = c(0, 60)))

res_cross <- synth_linked(
  tables     = list(patients = patients, admissions = adm),
  structures = list(patients = ~ id, admissions = ~ id / admission_id),
  keys       = list(patients = "id", admissions = c("id", "admission_id")),
  privacy    = dp_cross, seed = 1)
res_cross$privacy
```

The accounting now lists `admissions` as `cond. on parent` and the extra joints
raise the histogram count and the per-cell noise — the price of the added
structure, still inside the same $(\epsilon, \delta)$. Each child variable's
single strongest predictor may be a parent variable or another child variable
(under `dependence = "independent"` it is the best parent variable only), and a
three-level hierarchy chains it: a grandchild conditions on its already-conditioned
parent. Only the *immediate* parent conditions a child; deeper ancestors reach it
through the parent's synthesised values, as in Track A.

Only the *immediate* parent conditions a child; deeper ancestors reach it through
the parent's synthesised values, as in Track A. Constraints are refused, as
elsewhere under DP.

### Modelling a child table's rows over time

When a child table's rows are a *time series* within each parent unit — repeated
visits ordered by the child's own key index — `dp_control(longitudinal = TRUE)`
(or a vector of child-table names) models them as a within-unit **DP Markov
trajectory** instead of exchangeable records. The children-per-parent count model
doubles as the trajectory-length model; an initial-state model is measured over
each unit's first (earliest) child row; and a first-order transition matrix
$P(v_t \mid v_{t-1})$ is measured per variable over consecutive within-unit rows:

```{r linked-longi}
set.seed(1)
visits <- do.call(rbind, lapply(patients$id, function(pid) {
  k  <- 2L + rpois(1, 1.2)
  st <- character(k); st[1] <- sample(c("stable", "worse"), 1)
  for (i in 2:k) st[i] <- if (runif(1) < 0.85) st[i - 1]
                          else setdiff(c("stable", "worse"), st[i - 1])
  data.frame(id = pid, visit_num = seq_len(k),
             status = factor(st, levels = c("stable", "worse")))
}))

dp_longi <- dp_control(
  epsilon = 6, delta = 1e-6, mechanism = "gaussian",
  max_rows_per_person = c(visits = 6), longitudinal = "visits",
  domain = "public", bounds = list(age = c(18, 100)))

res_longi <- synth_linked(
  tables     = list(patients = patients, visits = visits),
  structures = list(patients = ~ id, visits = ~ id / visit_num),
  keys       = list(patients = "id", visits = c("id", "visit_num")),
  privacy    = dp_longi, seed = 1)
res_longi$privacy
```

The accounting lists `visits` as `DP Markov over rows`. The transition histograms
have person-sensitivity `path_cap[parent] * (branching_cap - 1)` while the
initial-state marginals sit at the parent path cap — heterogeneous, but folded into
the same exact $(\epsilon, \delta)$. Over-cap units are prefix-truncated in temporal
order so their consecutive pairs stay intact. This keeps a patient's status
*trending* across visits, the autocorrelation an exchangeable child would drop. A
longitudinally-modelled table is not simultaneously `cross_table`-conditioned — it
conditions on its own past, not the parent.

## What Track B does *not* do (yet)

- **Cross-table conditioning beyond the immediate parent** is not modelled;
  `cross_table = TRUE` conditions each child on its immediate parent only.
- **Cross-table conditioning *and* longitudinal transitions on the same child
  table** are mutually exclusive — a child is modelled one way or the other.
- Constraints (`rule()`) are refused under DP, because data-dependent rejection
  sampling would leak information the budget does not account for.
- `baseline` (columns held exactly constant within a person) and the
  higher-order / cross-variable transition model (`transition_order`,
  `transition_cross`) apply to a flat longitudinal `synth()` release; the
  longitudinally-modelled *linked* child table still uses the first-order,
  own-lag, all-time-varying model.

Within those limits the guarantee is exact and reported. For anything labelled
differentially private, keep the accounting record with the release. The
discretisation adds no unaccounted leakage in either rigorous mode — supply
public `bounds` (`domain = "public"`) to spend nothing on the domain, or let the
default `domain = "dp"` estimate the edges privately and charge them to the
budget.
