---
title: "Differentially private synthesis (Track B)"
output:
  litedown::html_format:
    meta:
      css: ["@default"]
vignette: >
  %\VignetteEngine{litedown::vignette}
  %\VignetteIndexEntry{Differentially private synthesis (Track B)}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
library(flexsynth)
```

## Two tracks, one interface

`flexsynth` has two engines. **Track A** (the default) maximises statistical
utility and carries *no* formal privacy guarantee — you judge the residual risk
with `diagnose()` and `disclosure_risk()`. **Track B** trades some utility for a
mathematically provable guarantee: **differential privacy (DP)**. You opt in by
passing a `dp_control()` to the `privacy` argument of `synth()`; nothing else
about the call changes.

```{r optin}
dp <- dp_control(epsilon = 1, delta = 1e-6, mechanism = "gaussian")
dp
```

Differential privacy bounds how much *any single individual* can change the
output distribution. The guarantee is stated as a pair
(\(\epsilon\), \(\delta\)): smaller values mean stronger privacy and more noise.
The default privacy unit here is **person-level** — the guarantee protects an
individual across *all* of their rows, not just a single record.

## A worked example

We use a small **synthetic** cardiac dataset (never real patient data): one row
per patient, with an age, sex, a smoking flag and a systolic blood pressure that
depends on both.

```{r data}
n <- 1500
real <- data.frame(
  id     = seq_len(n),
  age    = round(rnorm(n, 62, 11)),
  sex    = factor(sample(c("F", "M"), n, TRUE, prob = c(0.45, 0.55))),
  smoker = sample(c(FALSE, TRUE), n, TRUE, prob = c(0.7, 0.3))
)
real$sbp <- round(0.6 * real$age + ifelse(real$smoker, 8, 0) + rnorm(n, 90, 10))
head(real)
```

### Supply public bounds

The DP engine works over a **discrete grid**: numeric variables are cut into
bins. Where those bin edges come from matters. If you let the engine read them
from the data, that is a small data-dependent step that is *not* covered by the
(\(\epsilon\), \(\delta\)) accounting — the engine warns you when this happens.
For a clean, fully data-independent release, pass public ranges (from domain
knowledge, a codebook, or physiological limits) via `bounds`:

```{r bounds}
dp <- dp_control(
  epsilon    = 2,
  delta      = 1e-6,
  mechanism  = "gaussian",
  bounds     = list(age = c(18, 100), sbp = c(60, 240))
)
res <- synth(real, structure = ~ id, privacy = dp, seed = 1)
res
```

The result is an ordinary `synth_result`; `as.data.frame()` gives the synthetic
table, whose `id` is a fresh surrogate key.

```{r syn}
syn <- as.data.frame(res)
head(syn)
```

### The privacy accounting

Every DP result carries an accounting record describing exactly what was spent
and how the noise was calibrated. This is the object to keep with a governed
release.

```{r acct}
res$privacy
```

The engine measures a set of low-order marginals under a single, composed budget:
one-way marginals for every variable and (with the default
`dependence = "tree"`) the pairwise marginals needed to learn a Chow-Liu
dependency tree. The number of marginals, the per-cell noise scale, and — for the
Gaussian mechanism — the zero-concentrated-DP \(\rho\) are all reported.

## Choices that matter

**Mechanism.** `"laplace"` gives *pure* \(\epsilon\)-DP (set `delta = 0`);
`"gaussian"` gives approximate (\(\epsilon\), \(\delta\))-DP with tighter
composition when many marginals are measured (it needs `delta > 0`).

**Dependence.** `"tree"` (default) keeps first- and second-order structure by
fitting a Chow-Liu tree over the noisy pairwise marginals — no extra budget,
because the same measurements are reused for structure and parameters.
`"independent"` measures only one-way marginals: fewer queries, so less noise per
query, but cross-variable correlations are lost. The real `age`/`sbp`
correlation here is about `r round(cor(real$age, real$sbp), 2)`. Under DP the tree
recovers a substantial part of it while the independent model necessarily drives
it to zero:

```{r dependence}
b <- list(age = c(18, 100), sbp = c(60, 240))
dp_tree  <- dp_control(epsilon = 6, mechanism = "laplace", dependence = "tree",
                       bins = 8, bounds = b)
dp_indep <- dp_control(epsilon = 6, mechanism = "laplace", dependence = "independent",
                       bins = 8, bounds = b)
s_tree  <- as.data.frame(synth(real, ~ id, privacy = dp_tree,  seed = 1))
s_indep <- as.data.frame(synth(real, ~ id, privacy = dp_indep, seed = 1))
c(real  = cor(real$age, real$sbp),
  tree  = cor(s_tree$age,  s_tree$sbp),
  indep = cor(s_indep$age, s_indep$sbp))
```

Two things are visible in that vector: the tree keeps a clear positive
association (never the full strength — DP noise attenuates it), and the
independent model has no correlation at all by construction. The gap between the
tree and the real value is the honest price of the guarantee; a larger `epsilon`
or a coarser `bins` narrows it.

**Contribution bound.** At `unit = "person"` each person may contribute at most
`max_rows_per_person` rows (default `1`, appropriate for one-row-per-person
tables). If a person legitimately has several rows, set this from public domain
knowledge; rows beyond the cap are subsampled away before any budget is spent, so
the guarantee holds by construction.

**More budget, more fidelity.** Larger `epsilon` means less noise. The synthetic
marginals converge to the real ones as \(\epsilon\) grows:

```{r budget}
for (e in c(0.5, 2, 8)) {
  d <- dp_control(epsilon = e, mechanism = "laplace",
                  dependence = "independent",
                  bounds = list(age = c(18, 100), sbp = c(60, 240)))
  s <- as.data.frame(synth(real, ~ id, privacy = d, seed = 1))
  cat(sprintf("epsilon = %-3g  mean(sbp): real %.1f  syn %.1f\n",
              e, mean(real$sbp), mean(s$sbp)))
}
```

## What Track B does *not* do (yet)

- It produces a **flat** marginal release. Within-unit / longitudinal structure
  (the autoregressive temporal model of Track A) is **not** preserved under DP;
  a `structure = ~ id / visit` is treated as a flat table with a surrogate id.
- **Linked multi-table** DP (via `synth_linked()`) is not available yet.
- Constraints (`rule()`) are refused under DP, because data-dependent rejection
  sampling would leak information the budget does not account for.

Within those limits the guarantee is exact and reported. For anything labelled
differentially private, keep the accounting record with the release, and prefer
public `bounds` so the discretisation adds no unaccounted leakage.
