---
title: "Getting started with flexsynth"
output:
  litedown::html_format:
    meta:
      css: ["@default"]
vignette: >
  %\VignetteEngine{litedown::vignette}
  %\VignetteIndexEntry{Getting started with flexsynth}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
library(flexsynth)
```

## What flexsynth does

`flexsynth` generates synthetic data that reproduces the statistical structure
of a real dataset without copying its records. The default **Track A** engine is
a high-utility sequential synthesiser; it carries **no** formal privacy
guarantee, so the package ships honest utility and disclosure-risk diagnostics
so you can judge the trade-off yourself. (An opt-in differentially private
**Track B** is a later roadmap phase.)

## A first synthesis

We start with a plain single-table dataset — one row per patient.

```{r data}
n <- 300
real <- data.frame(
  id     = seq_len(n),
  age    = round(rnorm(n, 62, 11)),
  sex    = sample(c("F", "M"), n, replace = TRUE, prob = c(0.45, 0.55)),
  smoker = sample(c(FALSE, TRUE), n, replace = TRUE, prob = c(0.7, 0.3))
)
real$sbp <- round(0.6 * real$age + ifelse(real$smoker, 8, 0) + rnorm(n, 90, 10))
head(real)
```

`synth()` takes the data and a `structure` formula naming the unit identifier.
Here every row is its own unit, so the structure is just `~ id`.

```{r synth}
res <- synth(real, structure = ~ id, seed = 1)
res
syn <- as.data.frame(res)
head(syn)
```

The unit identifier is regenerated, and every other column is synthesised in
turn from a model fitted on the real data (the default `method = "cart"` draws
from the matching leaf of a regression / classification tree).

## Is the synthetic data any good? — `diagnose()`

`diagnose()` compares real and synthetic data on three views: per-variable
marginals (a Kolmogorov-Smirnov statistic for numeric variables, total-variation
distance for categorical ones), the correlation structure, and a general
**propensity (pMSE)** score — a model tries to tell real from synthetic records,
and a `ratio` near 1 means it cannot.

```{r diagnose}
d <- diagnose(real, res)
d
```

The `plot()` method overlays each marginal.

```{r diag-plot, fig.width = 7, fig.height = 5}
plot(d)
```

## How risky is it? — `disclosure_risk()`

Synthetic data is **not** anonymisation. `disclosure_risk()` reports how much the
synthetic data could leak about the real records: replicated uniques (real
records that are unique on the quasi-identifiers and reproduced exactly),
distance to closest record, and — given a holdout — a membership-inference
check. Pass the genuinely identifying columns as `quasi` (not surrogate keys
like the regenerated `id`).

```{r risk}
disclosure_risk(real, res, quasi = c("age", "sex", "smoker", "sbp"), seed = 1)
```

## Choosing a method

Methods live in an extensible registry. Built-ins include `sample`, `cart`,
`forest` (a bagged CART ensemble), `ctree` (via `partykit`), and the parametric
numeric methods `norm` and `normrank`.

```{r methods}
list_methods()
```

Set one method for all variables, or per variable via `synth_control()`:

```{r per-var}
ctrl <- synth_control(method = c(sbp = "norm", age = "cart"))
res2 <- synth(real, ~ id, tuning = ctrl, seed = 1)
```

You can register your own with `register_method()` — supply a `fit()` and a
`draw()`.

## Constraints

`rule()` declares a logical constraint the synthetic data must satisfy;
`synth()` enforces it by rejection sampling at the unit grain.

```{r rules}
res3 <- synth(real, ~ id,
              constraints = rule(sbp >= 80 & sbp <= 220),
              seed = 1)
range(as.data.frame(res3)$sbp)
```

## Tuning

`synth_control()` also exposes `smoothing` (kernel-smooth numeric draws) and
`predictor_matrix` (restrict which variables may predict each target). See
`?synth_control`.

Track A output must never be described as differentially private. Use the
diagnostics above to decide whether the utility / risk balance is acceptable for
your release.
