---
title: "Valid inference and attribute-disclosure control"
output:
  litedown::html_format:
    meta:
      css: ["@default"]
vignette: >
  %\VignetteEngine{litedown::vignette}
  %\VignetteIndexEntry{Valid inference and attribute-disclosure control}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
library(flexsynth)
```

## Why analysing one synthetic dataset is not enough

`synth()` regenerates every column from a model fitted on the real data, using a
freshly drawn skeleton — so its output is *fully synthetic* in the Reiter (2003)
sense. That randomness is real: run the synthesiser twice and your regression
coefficients move a little. An analyst who fits a model on a **single** synthetic
dataset and reports its ordinary standard errors is ignoring that extra
variability, and the standard errors come out too small — confidence intervals
that are too narrow, p-values that are too keen.

The fix is the reason `synth()` can emit `m > 1` datasets. Fit the analysis on
each, then combine the estimates with a rule that adds the between-synthesis
variability back in. `flexsynth` does this for you.

```{r data}
n <- 800
real <- data.frame(
  id  = seq_len(n),
  age = round(rnorm(n, 62, 11)),
  sex = sample(c("F", "M"), n, replace = TRUE, prob = c(0.45, 0.55))
)
real$sbp <- round(0.6 * real$age + ifelse(real$sex == "M", 5, 0) +
                    rnorm(n, 90, 10))

res <- synth(real, structure = ~ id, m = 10, seed = 1)
res
```

We asked for `m = 10` synthetic datasets. For pooled inference, more is better;
`m >= 5` is a sensible floor.

## `synth_glm()` — the common case

For a linear or generalised-linear model, `synth_glm()` fits it on every
synthetic dataset and pools the result in one call.

```{r glm}
pooled <- synth_glm(res, sbp ~ age + sex)
pooled
```

Compare that with the model fitted on the **real** data — the estimand we are
trying to recover:

```{r real-fit}
summary(lm(sbp ~ age + sex, data = real))$coefficients[, 1:2]
```

The pooled point estimates land on the real coefficients (the `age` slope near
0.6, the male offset near 5), and the pooled standard errors are honest about the
cost of synthesis rather than pretending a single draw was the truth.

By default the standard error targets the population the real data was drawn
from. If instead you want the standard error an analyst of the *original* sample
would have reported, set `population_inference = FALSE`.

## `pool_synth()` — any estimator

`synth_glm()` is a thin wrapper. For anything else — a quantile, a survival
model, a hand-rolled statistic — pass `pool_synth()` a function that runs on one
synthetic `data.frame` and returns either a fitted model (with `coef()` /
`vcov()` methods) or an explicit `list(estimate =, variance =)`.

```{r pool}
# A proportion and its variance, supplied directly.
pool_synth(res, function(d) {
  p <- mean(d$sex == "M")
  list(estimate = c(prop_male = p),
       variance = c(prop_male = p * (1 - p) / nrow(d)))
})
```

Two combining rules are available via `rule =`: the default `"synthpop"`
estimators of Raab, Nowok & Dibben (2016) — stable at small `m` and never
negative — and the classic `"reiter"` fully-synthetic estimator, which is
theoretically standard but needs `m >= 2` and can occasionally return a negative
variance (reported as `NA`).

## `compare_estimates()` — analysis-specific utility

`diagnose()` (see `vignette("getting-started")`) asks a general question: can a
model tell real from synthetic records? `compare_estimates()` asks the question
an analyst actually cares about: *does my regression come out the same?* It fits
the analysis on both sides — pooling the synthetic side automatically — and
reports, per term, the confidence-interval overlap (Karr et al. 2006; 1 means
identical intervals, 0 means they just touch, negative means disjoint) and the
standardised difference of the estimates.

```{r compare}
compare_estimates(real, res, function(d) lm(sbp ~ age + sex, data = d))
```

A mean overlap near 1 with small standardised differences is what you hope to
see: the synthetic data supports the same conclusions as the real data.

## Attribute disclosure — TCAP

Reproducing an analysis well is a *utility* win, but the same key-to-target
structure an analyst exploits is what an attacker exploits. `disclosure_risk()`
gains a `target =` argument for exactly this: the **Target Correct Attribution
Probability** (TCAP; Taub, Elliot et al.). An attacker who knows a real record's
quasi-identifier keys looks up the synthetic records sharing those keys and reads
off the conditional distribution of the sensitive `target`; TCAP is the mean
synthetic probability of the true value, and is compared with a marginal-only
attacker so the `lift` is the *excess* attribution the keys buy.

TCAP is meant for categorical keys and target, so we use a small categorical
dataset where a sensitive status depends on the quasi-identifiers.

```{r tcap-data}
m <- 800
clinic <- data.frame(
  region = sample(c("N", "S", "E", "W"), m, replace = TRUE),
  agecat = sample(c("40s", "50s", "60s", "70s"), m, replace = TRUE)
)
# hypertension risk rises steeply with age, with a small regional bump.
prob_htn <- c("40s" = 0.12, "50s" = 0.40, "60s" = 0.75, "70s" = 0.93)
reg_bump <- c("N" = 0, "S" = 0, "E" = 0.03, "W" = 0.05)
p <- pmin(0.98, prob_htn[clinic$agecat] + reg_bump[clinic$region])
clinic$hypertension <- ifelse(runif(m) < p, "yes", "no")
clinic$id <- seq_len(m)

cres <- synth(clinic, ~ id, m = 1, seed = 1)
```

Name the sensitive column as `target`; the remaining quasi-identifiers become the
attacker's keys.

```{r tcap}
disclosure_risk(clinic, cres,
                quasi  = c("region", "agecat", "hypertension"),
                target = "hypertension",
                seed   = 1)
```

Read the `Attribute disclosure` block: a `lift` well above zero means the
synthetic data lets an attacker attribute the true `hypertension` status
meaningfully better than the marginal rate alone — here because `agecat` carries
real signal about it. That is a governance signal, not a verdict: some attribute
association is the whole point of useful data. Weigh the lift against how
sensitive the target is and who receives the release. A `lift` near zero means
the keys add almost nothing beyond the margin.

## Where this fits

Track A output must never be described as differentially private; the tools here
help you decide whether its utility / disclosure balance is acceptable for a
given release. For a formal privacy guarantee instead, see
`vignette("differential-privacy")`. Note that pooled inference (`pool_synth()`,
`synth_glm()`) is a Track A tool — valid inference from a differentially private
release must additionally account for the DP noise and is out of scope here.
